---
title: "Box-Jenkins Forecasting Analysis: Global Price of Cocoa"
author: "Lucas Casimiro 20241796 / Lucas Christol 20241827 / Miguel Gomes 20241791 / Pedro Ferreira 20241735"
output: html_document
date: "2026-05-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message= FALSE)
```

# 1. Imports and data preparation

```{r packages}
library(fredr) 
library(tsibble) 
library(fable) 
library(feasts) 
library(dplyr) 
library(ggplot2)
library(lubridate)
```

```{r set-api-key}
fredr_set_key("14da34f1201fa46c838595fe42148d06")
```

```{r fetch-cocoa-data}
# Download dataset from FRED since 2000-01
raw_cocoa <- fredr(
  series_id = "PCOCOUSDM",
  observation_start = as.Date("2000-01-01"), 
  observation_end = Sys.Date(),
  frequency = "m"
)

# Convert to clean monthly tsibble
cocoa_ts <- raw_cocoa %>%
  mutate(Month = yearmonth(date)) %>%
  select(Month, value) %>%
  rename(Price = value) %>%
  as_tsibble(index = Month)
```

```{r split-data}
# Splitting train and test data, using last 2 years for testing
cocoa_train <- cocoa_ts %>% filter(Month <= max(Month) - 24)
cocoa_test <- cocoa_ts %>% filter(Month > max(Month) - 24)
```

*Write your introduction and context on cocoa price dynamics and dataset relevance here.*

# 2. TENTATIVE IDENTIFICATION PHASE

## 2.1 Visualize the Raw Time Series

```{r plot-raw}
cocoa_ts %>%
  autoplot(Price) +
  labs(
    title = "Global Price of Cocoa (PCOCOUSDM): 2000–2026",
    y = "USD per Metric Ton",
    x = "Year"
  ) +
  theme_minimal()
```

*Discuss major trends, price shocks, and any visible seasonality.*

## 2.2 Log Transformation for Variance Stabilization

```{r log-transform}
# Adding log-price column for later modeling
cocoa_ts <- cocoa_ts %>%
  mutate(log_price = log(Price))
```

*Briefly justify why log transformation is suitable given volatility and shocks.*

## 2.3 Differencing and ACF/PACF Analysis

```{r acf-pacf}
# Visualizing log-price, seasonal, and non-seasonal lags
cocoa_ts %>%
  gg_tsdisplay(log_price, plot_type = "partial")
```

```{r acf-pacf}
# Visualizing log-price, seasonal, and non-seasonal lags
cocoa_ts %>%
  gg_tsdisplay(difference(log_price), plot_type = "partial")
```

*Interpret the ACF and PACF patterns you see here (indicate likely values for d and D).*

## 2.4 Unit Root Tests (Mathematical Differencing Justification)

```{r unit-root}
# Computing required differencing orders
ndiffs <- cocoa_ts %>%
  features(log_price, features = unitroot_ndiffs)

nsdiffs <- cocoa_ts %>%
  features(log_price, features = unitroot_nsdiffs)

ndiffs
nsdiffs
```

*State which differences you will use and why, based on the outputs above.*

# 3. MODEL ESTIMATION PHASE

## 3.1 Fit Candidate SARIMA Models

```{r fit-models}
# Manual SARIMA: adjust pdq/PDQ per identification above
fit_manual <- cocoa_ts %>%
  model(SARIMA_manual = ARIMA(log_price ~ pdq(1,1,1) + PDQ(1,1,1)[12]))

# Automatic SARIMA search as a benchmark
fit_auto <- cocoa_ts %>%
  model(SARIMA_auto = ARIMA(log_price))
```

*Justify your manual candidate orders using previous diagnostics. Note what the auto model selects.*

## 3.2 Compare Model Fit (AICc Table)

```{r compare-models}
library(broom)
model_stats <- bind_rows(
  glance(fit_manual) %>% mutate(model = "Manual SARIMA"),
  glance(fit_auto) %>% mutate(model = "Auto SARIMA")
)

knitr::kable(
  model_stats %>% select(model, aic, aicc, bic, log_lik),
  digits = 2, caption = "SARIMA Model Comparison"
)
```

*Interpret fit criteria and state which model is preferred.*

# 4. DIAGNOSTIC CHECKING PHASE

## 4.1 Residual Plots (White Noise Assessment)

```{r residual-plots}
# Use best model from above (replace SARIMA_auto if manual is better)
best_fit <- fit_auto

best_fit %>% gg_tsresiduals()
```

*Examine and comment on white noise, autocorrelation, and outliers in residuals.*

## 4.2 Ljung-Box Test for Residual Independence

```{r ljung-box}
# Ljung-Box test
# Adjust dof to match number of estimated parameters in your selected model
best_fit %>%
  augment() %>%
  features(.innov, ljung_box, lag = 24, dof = 6)
```

*Interpret whether autocorrelation appears in the residuals (p-values).*

# 5. FORECASTING PHASE

## 5.1 24-Month (2-Year) Ahead Ex-ante Forecast

```{r forecast-24mo}
# Forecast next 2 years (24 months)
forecast_24mo <- best_fit %>%
  forecast(h = "2 years")

forecast_24mo %>%
  autoplot(cocoa_ts, level = c(80, 95)) +
  labs(
    title = "24-Month Forecast: Global Price of Cocoa (log-scale)",
    y = "Log(USD per Metric Ton)",
    x = "Year"
  ) +
  theme_minimal()
```

*Interpret the forecast levels, trend, and confidence intervals for stakeholders.*

# 6. FINAL SUBMISSION: DATA EXPORT

```{r save-tsibble}
# Saving the cleaned cocoa tsibble for reproducibility
save(cocoa_ts, file = "data_group_10.RData")
```

## SESSION INFO

```{r session-info, echo=FALSE}
sessionInfo()
```
