---
title: "Box-Jenkins Forecasting Analysis: Global Price of Cocoa"
author: "Lucas Casimiro 20241796 / Lucas Christol 20241827 / Miguel Gomes 20241791 / Pedro Ferreira 20241735"
output: html_document
date: "2026-05-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message= FALSE)
```

# 1. Introduction

Global cocoa prices have been an interesting topic as of late, due to severe weather conditions and diseases affecting crops in West Africa, one of the main supply areas. We utilize the Global Price of Cocoa (PCOCOUSDM) monthly dataset from FRED to analyze the dynamics involved, while employing the Box-Jenkins methodology, a robust approach for time series analysis and forecasting.

```{r packages}
library(fredr) 
library(tsibble) 
library(fable) 
library(feasts) 
library(dplyr) 
library(ggplot2)
library(lubridate)

# Using API key for access to dataset, from library fredr
fredr_set_key("14da34f1201fa46c838595fe42148d06")

# Filter dataset from FRED since 2000
raw_cocoa <- fredr(
  series_id = "PCOCOUSDM",
  observation_start = as.Date("2000-01-01"), 
  observation_end = Sys.Date(),
  frequency = "m"
)

# Convert to tsibble
cocoa_ts <- raw_cocoa %>%
  mutate(Month = yearmonth(date)) %>%
  select(Month, value) %>%
  rename(Price = value) %>%
  as_tsibble(index = Month)

```

```{r plot-raw}
cocoa_ts %>%
  autoplot(Price) +
  labs(title = "Global Price of Cocoa: 2000–2026",
    y = "USD per Metric Ton",
    x = "Year") + theme_minimal()
```

Historically, global cocoa prices fluctuated around 2000\$ and 3000\$, exhibiting long-term stability. However, we can clearly see from the raw data when the complications mentioned started (late 2023), and the price skyrockets to values over 8000\$, even reaching 10000\$. A slight correction is noticed beginning last year, with closer to normal values recently.

```{r split-data}
# Splitting train and test data, using last 2 years (24 months) for testing
cocoa_train <- cocoa_ts %>% 
  filter(Month <= max(Month) - 24)
cocoa_test <- cocoa_ts %>% 
  filter(Month > max(Month) - 24)
```

# 2. TENTATIVE IDENTIFICATION PHASE

## 2.2 Log Transformation for Variance Stabilization

```{r log-transform}
# Adding log-price column for later modeling
cocoa_ts <- cocoa_ts %>%
  mutate(log_price = log(Price))
```

*Briefly justify why log transformation is suitable given volatility and shocks.*

## 2.3 Differencing and ACF/PACF Analysis

```{r acf-pacf}
# Visualizing log-price, seasonal, and non-seasonal lags
cocoa_ts %>%
  gg_tsdisplay(log_price, plot_type = "partial")
```

```{r acf-pacf}
# Visualizing log-price, seasonal, and non-seasonal lags
cocoa_ts %>%
  gg_tsdisplay(difference(log_price), plot_type = "partial")
```

*Interpret the ACF and PACF patterns you see here (indicate likely values for d and D).*

## 2.4 Unit Root Tests (Mathematical Differencing Justification)

```{r unit-root}
# Computing required differencing orders
ndiffs <- cocoa_ts %>%
  features(log_price, features = unitroot_ndiffs)

nsdiffs <- cocoa_ts %>%
  features(log_price, features = unitroot_nsdiffs)

ndiffs
nsdiffs
```

*State which differences you will use and why, based on the outputs above.*

# 3. MODEL ESTIMATION PHASE

## 3.1 Fit Candidate SARIMA Models

```{r fit-models}
# Manual SARIMA: adjust pdq/PDQ per identification above
fit_manual <- cocoa_ts %>%
  model(SARIMA_manual = ARIMA(log_price ~ pdq(1,1,1) + PDQ(1,1,1)[12]))

# Automatic SARIMA search as a benchmark
fit_auto <- cocoa_ts %>%
  model(SARIMA_auto = ARIMA(log_price))
```

*Justify your manual candidate orders using previous diagnostics. Note what the auto model selects.*

## 3.2 Compare Model Fit (AICc Table)

```{r compare-models}
library(broom)
model_stats <- bind_rows(
  glance(fit_manual) %>% mutate(model = "Manual SARIMA"),
  glance(fit_auto) %>% mutate(model = "Auto SARIMA")
)

knitr::kable(
  model_stats %>% select(model, aic, aicc, bic, log_lik),
  digits = 2, caption = "SARIMA Model Comparison"
)
```

*Interpret fit criteria and state which model is preferred.*

# 4. DIAGNOSTIC CHECKING PHASE

## 4.1 Residual Plots (White Noise Assessment)

```{r residual-plots}
# Use best model from above (replace SARIMA_auto if manual is better)
best_fit <- fit_auto

best_fit %>% gg_tsresiduals()
```

*Examine and comment on white noise, autocorrelation, and outliers in residuals.*

## 4.2 Ljung-Box Test for Residual Independence

```{r ljung-box}
# Ljung-Box test
# Adjust dof to match number of estimated parameters in your selected model
best_fit %>%
  augment() %>%
  features(.innov, ljung_box, lag = 24, dof = 6)
```

*Interpret whether autocorrelation appears in the residuals (p-values).*

# 5. FORECASTING PHASE

## 5.1 24-Month (2-Year) Ahead Ex-ante Forecast

```{r forecast-24mo}
# Forecast next 2 years (24 months)
forecast_24mo <- best_fit %>%
  forecast(h = "2 years")

forecast_24mo %>%
  autoplot(cocoa_ts, level = c(80, 95)) +
  labs(
    title = "24-Month Forecast: Global Price of Cocoa (log-scale)",
    y = "Log(USD per Metric Ton)",
    x = "Year"
  ) +
  theme_minimal()
```

*Interpret the forecast levels, trend, and confidence intervals for stakeholders.*

# 6. FINAL SUBMISSION: DATA EXPORT

```{r save-tsibble}
# Saving the cleaned cocoa tsibble for reproducibility
save(cocoa_ts, file = "data_group_13.RData")
```

## SESSION INFO

```{r session-info, echo=FALSE}
sessionInfo()
```
