---
title: "Box-Jenkins Forecasting Analysis: Global Price of Cocoa"
author: "Lucas Casimiro 20241796 / Lucas Christol 20241827 / Miguel Gomes 20241791 / Pedro Ferreira 20241735"
output: html_document
date: "2026-05-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message= FALSE)
```

# 1. Introduction

Global cocoa prices have been an interesting topic as of late, due to severe weather conditions and diseases affecting crops in West Africa, one of the main supply areas. We utilize the Global Price of Cocoa (PCOCOUSDM) monthly dataset from FRED to analyze the dynamics involved, while employing the Box-Jenkins methodology, a robust approach for time series analysis and forecasting.

```{r packages}
library(fredr) 
library(tsibble) 
library(fable) 
library(feasts) 
library(dplyr) 
library(ggplot2)
library(lubridate)

# Using API key for access to dataset, from library fredr
fredr_set_key("14da34f1201fa46c838595fe42148d06")

# Filter dataset from FRED since 2000
raw_cocoa <- fredr(
  series_id = "PCOCOUSDM",
  observation_start = as.Date("2000-01-01"), 
  observation_end = Sys.Date(),
  frequency = "m"
)

# Convert to tsibble
cocoa_ts <- raw_cocoa %>%
  mutate(Month = yearmonth(date)) %>%
  select(Month, value) %>%
  rename(Price = value) %>%
  as_tsibble(index = Month)

```

```{r plot-raw}
cocoa_ts %>%
  autoplot(Price) +
  labs(title = "Global Price of Cocoa: 2000–2026",
    y = "USD per Metric Ton",
    x = "Year") + theme_minimal()
```

Historically, global cocoa prices fluctuated around 2000\$ and 3000\$, exhibiting long-term stability. However, we can clearly see from the raw data when the complications mentioned started (late 2023), and the price skyrockets to values over 8000\$, even reaching 10000\$. A slight correction is noticed beginning last year, with closer to normal values recently.

```{r split-data}
# Splitting train and test data, using last 2 years (24 months) for testing
cocoa_train <- cocoa_ts %>% 
  filter(Month <= max(Month) - 24)
cocoa_test <- cocoa_ts %>% 
  filter(Month > max(Month) - 24)
```

The dataset is partitioned into a training set (from 2000 through mid-2024) and a testing set including the last 2 years. This testing window is strategic given the unusual spike as of recent years, as it allows our model to be stress-tested in extreme market conditions.

While keeping a massive training dataset with the long-term stability and the beginning of the cocoa price crisis, the testing window forces us to evaluate whether a traditional SARIMA structure can adapt to these conditions.

# 2. TENTATIVE IDENTIFICATION PHASE

## Log Transformation

```{r log-transform}
# Adding log-price column for later modeling
cocoa_train <- cocoa_train %>%
  mutate(log_price = log(Price))
```

Visual inspection of the raw time series reveals heteroscedacity. When the price of cocoa increases, the variance increases. To stabilize these changes, we apply a log transformation before conducting stationary tests or differencing.

## Differencing and ACF/PACF Analysis

```{r acf-pacf}
# Visualizing log-price, seasonal, and non-seasonal lags
cocoa_train %>%
  gg_tsdisplay(log_price, plot_type = "partial")
```

Before differencing, we can see the ACF shows a very slow, linear decay across all lags, and the PACF has a massive spike at lag 1 and white noise after. These plots are characteristic of a non-stationary random walk, which confirms the presence of a unit root, and the need for differencing.

```{r acf-pacf-stationary}
# Same visualization after differencing
cocoa_train %>%
  gg_tsdisplay(difference(log_price), plot_type = "partial")
```

After taking the first difference (d=1), the series now looks stationary. The ACF plot shows a significant spike at lag 1 followed by a cut-off, same as the PACF. Because both function exhibit the same behavior, visually distinguishing between a pure AR or a pure MA model is ambiguous. Therefore, we identify ARIMA(1,1,0) and ARIMA(0,1,1) as strong baseline models, while also including a ARIMA(1,1,1) candidate to capture potential overlapping.

## Unit Root Testing

```{r unit-root}
# Computing required differencing orders
ndiffs <- cocoa_train %>%
  features(log_price, features = unitroot_ndiffs)

nsdiffs <- cocoa_train %>%
  features(log_price, features = unitroot_nsdiffs)

ndiffs
nsdiffs
```

To mathematically confirm our visual analysis regarding stationarity above, unit root testing was used (testing for both seasonal and non-seasonal differences). The ndiffs test indicated that one non-seasonal difference is necessary (confirming previous observation). The nsdiffs test indicated that no seasonal differencing is needed. This means our candidate models will proceed with d=1 and D=0.

## Information Criteria for Model Identification

Information Criteria is used as a supplementary guide to our ACF/PACF analysis, with the intention of helping in the decision between the candidate models. The candidate models are fitted and their information criteria is showed in a table.

Given our moderate sample size, we prioritize AICc (and AIC) over BIC, as AIC-based criteria perform better in smaller samples, while BIC can become unreliable and select overparameterized models in this case.

```{r fit-models}
# Fitting candidate SARIMA models from previous analysis
fit_models <- cocoa_train %>%
  model(
    # MA(1) non-seasonal
    SARIMA_011_001 = ARIMA(log_price ~ pdq(0,1,1) + PDQ(0,0,1, period=12)),
    
    # AR(1) non-seasonal
    SARIMA_110_100 = ARIMA(log_price ~ pdq(1,1,0) + PDQ(1,0,0, period=12)),
    
    # ARMA(1,1) + seasonal MA term
    SARIMA_111_001 = ARIMA(log_price ~ pdq(1,1,1) + PDQ(0,0,1, period=12)),
        
    # ARMA(1,1) + seasonal AR term
    SARIMA_111_100 = ARIMA(log_price ~ pdq(1,1,1) + PDQ(1,0,0, period=12)),
    
    # Automatic search to serve as a benchmark   
    SARIMA_auto = ARIMA(log_price))

# Displaying the information criteria table for identification selection 
knitr::kable(
  glance(fit_models) %>% 
    select(.model, AIC, AICc, BIC) %>%
    arrange(AICc),
  digits = 2, caption = "SARIMA Model Comparison"
)
```

Interestingly, our results don't reach a perfect consensus. SARIMA_auto minimizes both AICc and AIC. However, the BIC defends the simpler manual configurations SARIMA_011_001 and SARIMA_110_100.

Following the principle defined, we proceed with SARIMA_auto into the estimation and diagnostic phases, checking if its additional parameter complexity is truly justified.

# 3. MODEL ESTIMATION PHASE

With our tentatively identified model, we move to the formal estimation phase using Maximum-Likelihood Estimation. Also, we create a table to compare the log-likelihood for every model.

```{r estimate-chosen}
best_model <- fit_models %>% select(SARIMA_auto)

# Table for log-likelihood comparison
knitr::kable(
  glance(fit_models) %>% 
    select(.model, log_lik) %>%
    arrange(log_lik),
  digits = 2, caption = "Log-Likelihood Comparison"
)

best_model %>%
  report()
  
```

We can see that the highest raw log-likelihood value belongs to the SARIMA_auto model, the one we chose, which indicates a better mathematical fit.

The report output shows that SARIMA_auto achieved this superior fit by estimating a ARIMA(2,1,2)(1,0,0)[12] w/ drift structure. This validates our unit roots testing results, with d=1 and D=0, while uncovering higher terms and an upward trend that were omitted during visual inspection of the plots.

The estimated parameters for this model are $\phi_1 = -0.8521$, $\phi_2 = -0.5445$, $\theta_1 = 1.1223$, $\theta_2 = 0.6808$, and a seasonal autoregressive term $\Phi_1 = 0.0065$. While the non-seasonal parameters are significant given their standard error, the seasonal parameter has a standard error that vastly exceeds its own estimate, indicating that said component is statistically insignificant.

However, the estimation values are completely invalid if the candidate model has structural flaws or over-parametrization. This means we can only reach a conclusion after verifying if this model is valid in our diagnostic checking phase.

# 4. DIAGNOSTIC CHECKING PHASE

## Residual Plots (White Noise Assessment)

```{r residual-plots}
# Plotting the residuals from the model chosen
best_model %>% gg_tsresiduals()
```

*Examine and comment on white noise, autocorrelation, and outliers in residuals.*

## 4.2 Ljung-Box Test for Residual Independence

```{r ljung-box}
# lag = 24 for monthly data, dof = 5 (ARIMA(2,1,2)(1,0,0)[12], 2+1+2=5)
best_model %>%
  augment() %>%
  features(.innov, ljung_box, lag = 24, dof = 5)
```

*Interpret whether autocorrelation appears in the residuals (p-values).*

# 5. FORECASTING PHASE

## 24-Month (2-Year) Forecast

```{r forecast-24mo}
# Forecast next 2 years 
forecast_24mo <- best_model %>%
  forecast(h = "2 years")

forecast_24mo %>%
  autoplot(cocoa_ts, level = c(80, 95)) +
  labs(
    title = "24-Month Forecast: Global Price of Cocoa (log-scale)",
    y = "Log(USD per Metric Ton)",
    x = "Year"
  ) +
  theme_minimal()
```

*Interpret the forecast levels, trend, and confidence intervals for stakeholders.*

# 6. FINAL SUBMISSION: DATA EXPORT

```{r save-tsibble}
# Saving the cleaned cocoa tsibble for reproducibility
save(cocoa_ts, file = "data_group_13.RData")
```

## SESSION INFO

```{r session-info, echo=FALSE}
sessionInfo()
```
